Apparent preferences for cognitive effort fade when multiple forms of effort

### 2021-03-04 Cost_manuscript.Rmd
revisions from dissertation

\begin{align}
  \text{Model}_\text{biased} = R_i - (\gamma + mC_i) H_i
\end{align}

@@

\begin{align}
  mC_i = C w_i + c_{demand} (1 - w_i)
\end{align}

@@

\begin{align}
  \text{Model}_\text{adaptive + biased} = R_i - (\gamma_i + mC_i) H_i
\end{align}

@@

@@

  Next, we tested the hypothesis that foragers would display consistent and stable choices. Participants were deemed consistent if they did not quit after engaging in a trial. To examine this, we plotted survival curves of trial quits across all participants (Figure 5, top-left), with black crosses signaling either completed trials (at 2 s, 10 s, and 14 s) or forced travels. These curves show that most decisions to quit happened within the first second into the handling time, and participants rarely quit during the handling time. Physical and easy groups showed slightly lagged responses because the experiment allowed a one-second grace period for participants to begin gripping (marked by the dashed line). This resulted in some individuals opting to wait for that second to indicate an offer rejection (the censor at 4s in the physical group was a completion during a 2 s handling trial that was incorrectly recorded as 4 s by the experimental code). Plotting the quitting time distributions across blocks for the first second of the handling time (ECDFs on Figure 5, top-right) showed that participants in the wait and cognitive groups responded faster as the experiment progressed. These response time observations suggested clarity in participant choices

@@

  Analyses from the previous experiment were repurposed to examine cost differences within individuals. We started by testing the hypothesis that participants would successfully integrate the statistics of the environment. A mixed-effects logistic regression once again showed that larger reward amounts significantly increased acceptance probabilities ($\beta$ = `r historyCoeffs_wth["Offer", "Estimate"]`, SE = `r historyCoeffs_wth["Offer", "Std. Error"]`, *p* `r report_p(historyCoeffs_wth["Offer", "Pr(>|z|)"])`), and that participants became more selective if they had recently observed large offers ($\beta$ = `r historyCoeffs_wth["recentOffers", "Estimate"]`, SE = `r historyCoeffs_wth["recentOffers", "Std. Error"]`, *p* `r report_p(historyCoeffs_wth["recentOffers", "Pr(>|z|)"])`). Model comparisons suggested that participants were influenced by the `r historyIndex_wth + 1` most recent rewards (AIC = `r round(AIC(historySize_wth[[historyIndex_wth]]), digits = 2)`). However, this time participants were unaffected by recent reward misses (*p* > 0.05).
  As before, a linear model predicting earnings showed that participants who completed too few or too many trials earned the least, in line with the experimental design (Figure 7A; general linear model with quadratic term, F = `r round(summary(lmEarn)$fstatistic[1], digits = 2)`, $\beta$ = `r round(summary(lmEarn)$coefficients[3,1], digits= 2)`, SE = `r round(summary(lmEarn)$coefficients[3, 2], digits = 2)`, $R^2$ = `r round(summary(lmEarn)$r.squared, digits = 2)`, *p* `r report_p(summary(lmEarn)$coefficients[3, 4])`). When assessing the optimality of acceptances for 4 and 8 cents (99% of 20 cent offers were correctly accepted), chi-squared tests against the optimal rate of 50% showed that participants under-accepted physical effort trials (proportion accepted = `r round(summary_costvsreward["Physical", "probability"], digits = 2)`, $\chi^2$ = `r round(summary_costvsreward["Physical", "chisquared"], digits = 2)`, *p* `r report_p(summary_costvsreward["Physical", "pval"])`; all other *p* > 0.05). Despite this deviation, participants earned a similar total amount per cost on average (mean Cognitive = `r filter(earnings_within, Cost == "Cognitive")$mEarn`, SD = `r filter(earnings_within, Cost == "Cognitive")$sdEarn`; mean Wait-C = `r filter(earnings_within, Cost == "Wait-C")$mEarn`, SD = `r filter(earnings_within, Cost == "Wait-C")$sdEarn`; mean Physical = `r filter(earnings_within, Cost == "Physical")$mEarn`, SD = `r filter(earnings_within, Cost == "Physical")$sdEarn`; mean Wait-P = `r filter(earnings_within, Cost == "Wait-P")$mEarn`, SD = `r filter(earnings_within, Cost == "Wait-P")$sdEarn`). These results convey that participants in both experiments were similarly influenced by the characteristics of the environment, although participants in the present experiment gravitated more towards optimality.

@@

@@

  However, against our predictions, participants accepted significantly fewer trials on the second half of the experiment, regardless of cost (paired permutations; Cognitive: mean~pre~ = `r filter(df_mean, Cost == "Cognitive")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Cognitive")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Cognitive")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Cognitive")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Cognitive"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Cognitive"]], digits = 2)`; Physical: mean~pre~ = `r filter(df_mean, Cost == "Physical")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Physical")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Physical")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Physical")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Physical"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Physical"]], digits = 2)`; Wait-C:  mean~pre~ = `r filter(df_mean, Cost == "Wait-C")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Wait-C")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Wait-C")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Wait-C")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Wait-C"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Wait-C"]], digits = 2)`; Wait-P: mean~pre~ = `r filter(df_mean, Cost == "Wait-P")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Wait-P")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Wait-P")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Wait-P")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Wait-P"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Wait-P"]], digits = 2)`; Figure 9). The reduction in acceptances was not driven by a performance decline, as the difference in acceptance rates was not significantly correlated with the proportion of forced travels across individuals (Pearson's r = `r round(corDiff, digits = 2)`, *p* > 0.05). Notably, the pattern of choices from the first half of the experiment resembled that from the matching handling/travel block from the between-subject experiment (Figure 4A). Mixed effects logistic regressions performed on each half separately confirmed that the relative preference for cognitive effort seen previously was preserved at the beginning of the experiment (Figure 9, top-right, lower triangle), but that it disappeared over time amid an overall decline in acceptance rates as participants gained experience with all three conditions (upper triangle). This fading preference for cognitive effort was observed regardless of which block participants experienced first, even though its magnitude appeared to be reduced for those who experienced the physical block first (Figure 9, bottom)

>>>>

\begin{align}
  \text{Model}_\text{biased} = R_i - (\gamma + C_i) H_i
\end{align}

@@

\begin{align}
  C_i = c_\text{global} w_i + c_\text{demand} (1 - w_i)
\end{align}

@@

\begin{align}
  \text{Model}_\text{adaptive + biased} = R_i - (\gamma_i + C_i) H_i
\end{align}

@@

# actual proportion of quits that happened after 1 s (i.e. after the grace period for physical)
handlingQuits_btw <- dataBtw %>%
  filter(rawChoice == 0) %>%
  summarize(r = round(mean(RT > 1.05), digits = 2)) # to account for some measurement error for the grace perior during grip

@@

  Next, we tested the hypothesis that foragers would display consistent and stable choices. Participants were deemed consistent if they did not quit after engaging in a trial. To examine this, we plotted survival curves of trial quits across all participants (Figure 5, top-left), with black crosses signaling either completed trials (at 2 s, 10 s, and 14 s) or forced travels. These curves show that most decisions to quit happened within the first second into the handling time, and participants rarely quit afterwards (proportion of quits after 1 s = `r  handlingQuits_btw$r`). Physical and easy groups showed slightly lagged responses because the experiment allowed a one-second grace period for participants to begin gripping (marked by the dashed line). This resulted in some individuals opting to wait for that second to indicate an offer rejection (the censor at 4s in the physical group was a completion during a 2 s handling trial that was incorrectly recorded as 4 s by the experimental code). Plotting the quitting time distributions across blocks for the first second of the handling time (ECDFs on Figure 5, top-right) showed that participants in the wait and cognitive groups responded faster as the experiment progressed. These response time observations suggested clarity in participant choices

@@

  Analyses from the previous experiment were repurposed to examine cost differences within individuals. We started by testing the hypothesis that participants would successfully integrate the statistics of the environment. A mixed-effects logistic regression once again showed that larger reward amounts significantly increased acceptance probabilities ($\beta$ = `r historyCoeffs_wth["Offer", "Estimate"]`, SE = `r historyCoeffs_wth["Offer", "Std. Error"]`, *p* `r report_p(historyCoeffs_wth["Offer", "Pr(>|z|)"])`), and that participants became more selective if they had recently observed large offers ($\beta$ = `r historyCoeffs_wth["recentOffers", "Estimate"]`, SE = `r historyCoeffs_wth["recentOffers", "Std. Error"]`, *p* `r report_p(historyCoeffs_wth["recentOffers", "Pr(>|z|)"])`). Model comparisons suggested that participants were influenced by the `r historyIndex_wth + 1` most recent rewards (AIC = `r round(AIC(historySize_wth[[historyIndex_wth]]), digits = 2)`). However, this time participants were unaffected by recent reward misses (*p* `r report_p(historyCoeffs_wth["recentTravels", "Pr(>|z|)"])`). As before, a linear model predicting earnings showed that participants who completed too few or too many trials earned the least, in line with the experimental design (Figure 7A; general linear model with quadratic term, F = `r round(summary(lmEarn)$fstatistic[1], digits = 2)`, $\beta$ = `r round(summary(lmEarn)$coefficients[3,1], digits= 2)`, SE = `r round(summary(lmEarn)$coefficients[3, 2], digits = 2)`, $R^2$ = `r round(summary(lmEarn)$r.squared, digits = 2)`, *p* `r report_p(summary(lmEarn)$coefficients[3, 4])`). When assessing the optimality of acceptances for 4 and 8 cents (99% of 20 cent offers were correctly accepted), chi-squared tests against the optimal rate of 50% showed that participants under-accepted physical effort trials (proportion accepted = `r round(summary_costvsreward["Physical", "probability"], digits = 2)`, $\chi^2$ = `r round(summary_costvsreward["Physical", "chisquared"], digits = 2)`, *p* `r report_p(summary_costvsreward["Physical", "pval"])`; all other *p* > 0.05). Despite this deviation, participants earned a similar total amount per cost on average (mean Cognitive = `r filter(earnings_within, Cost == "Cognitive")$mEarn`, SD = `r filter(earnings_within, Cost == "Cognitive")$sdEarn`; mean Wait-C = `r filter(earnings_within, Cost == "Wait-C")$mEarn`, SD = `r filter(earnings_within, Cost == "Wait-C")$sdEarn`; mean Physical = `r filter(earnings_within, Cost == "Physical")$mEarn`, SD = `r filter(earnings_within, Cost == "Physical")$sdEarn`; mean Wait-P = `r filter(earnings_within, Cost == "Wait-P")$mEarn`, SD = `r filter(earnings_within, Cost == "Wait-P")$sdEarn`). These results convey that participants in both experiments were similarly influenced by the characteristics of the environment, although participants in the present experiment gravitated more towards optimality.

@@

# no differences in forced travels across halves
temp <- dataWth %>%
  filter(!Cost %in% c("Wait-C", "Wait-P")) %>%
  group_by(SubjID,  Half) %>%
  summarise(propFails = mean(rawChoice == 2)) %>%
  spread(Half, propFails)

failDiffs_wth <- permute(temp$Half_1, temp$Half_2, paired = T)

@@

  However, against our predictions, participants accepted significantly fewer trials on the second half of the experiment, regardless of cost (paired permutations; Cognitive: mean~pre~ = `r filter(df_mean, Cost == "Cognitive")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Cognitive")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Cognitive")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Cognitive")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Cognitive"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Cognitive"]], digits = 2)`; Physical: mean~pre~ = `r filter(df_mean, Cost == "Physical")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Physical")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Physical")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Physical")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Physical"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Physical"]], digits = 2)`; Wait-C:  mean~pre~ = `r filter(df_mean, Cost == "Wait-C")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Wait-C")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Wait-C")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Wait-C")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Wait-C"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Wait-C"]], digits = 2)`; Wait-P: mean~pre~ = `r filter(df_mean, Cost == "Wait-P")$propAccept_1`, SD~pre~ = `r filter(df_mean, Cost == "Wait-P")$sdAccept_1`, mean~post~ = `r filter(df_mean, Cost == "Wait-P")$propAccept_2`, SD~post~ = `r filter(df_mean, Cost == "Wait-P")$sdAccept_2`, *p* `r report_p(prepost_wth$Perms[["Wait-P"]])`, Cohen's D = `r round(prepost_wth$CohenD[["Wait-P"]], digits = 2)`; Figure 9). The reduction in acceptances was not driven by a performance decline, forced travels were not significantly different in the first and second half of the experiment (paired permutation with 5000 iterations, *p* `r report_p(failDiffs_wth$Pval)`). Notably, the pattern of choices from the first half of the experiment resembled that from the matching handling/travel block from the between-subject experiment (Figure 4A). Mixed effects logistic regressions performed on each half separately confirmed that the relative preference for cognitive effort seen previously was preserved at the beginning of the experiment (Figure 9, top-right, lower triangle), but that it disappeared over time amid an overall decline in acceptance rates as participants gained experience with all three conditions (upper triangle). This fading preference for cognitive effort was observed regardless of which block participants experienced first, even though its magnitude appeared to be reduced for those who experienced the physical block first (Figure 9, bottom)
